A Framework for Reproducible Data Science: Managing Environments and Pipelines with Nix, rix, and rixpress

### 2025-10-16 paper.qmd
Update paper.qmd

---
title: "Reproducible development environments with rix"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  In order create an analysis that is reproducible, it is not enough to write
  clean code and document it well. One must also make sure to clearly list all
  the dependencies of the analysis and ideally provide an easy way to install
  them. There are several tools that can be used to list dependencies and to
  make them easily installable by someone that wishes to reproduce a study, such
  as [Docker]{.proglang}, a containerization solution. This paper will present
  another tool that can be used to make projects reproducible, the [Nix]{.pkg}
  package manager, and an [R]{.proglang} package called [rix]{.pkg} that makes
  using [Nix]{.pkg} much simpler for users of [R]{.proglang}.

@@

A final, and optional, step towards the gold standard is to use a build
automation tool such as [Make]{.pkg} to run the whole analysis when the
container is executed, instead of executing scripts. Build automation tools make
it easier to run arbitrary code in a series of well-defined steps.
[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
automation tool.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.pkg} for all the other software,
including the programming language that was used. Ideally, a build automation
tool such as [Make]{.pkg} should be added to the list. It should also be noted
that these requirements are not limited to the [R]{.proglang} programming
language. A very similar approach should be taken if one uses
[Python]{.proglang}, or any other programming language. Furthermore, the
previous discussion also implies that for an analysis to actually be
reproducible at all, it must be made using free and open source software, as
proprietary software cannot be freely shared with others that wish to reproduce
a study, and there's no guarantee that others will have access to the
proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.pkg} package
manager, which is available for all major operating systems and focuses on
installing and building software in a reproducible manner. This means that this
single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg} and even a
build automation tool such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
to [R]{.proglang} programmers, we developed the [rix]{.pkg} package which we
will present in this article.

@@

>>>>

---
title: "A Framework for Reproducible Data Science: Managing Environments and Pipelines with Nix, rix, and rixpress"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Bruno Rodrigues 
    affiliations:
      - name: Ministry of Research and Higher education, Luxembourg 
        department: Department of Statistics
        address: 18, Montée de la Pétrusse
        city: Luxembourg 
        country: Luxembourg 
        postal-code: 2327
      - Journal of Statistical Software
    orcid: 0000-0002-3211-3689
    email: bruno@brodrigues.co
    url: https://www.brodrigues.co 
  - name: Philipp Baumann
    affiliations:
      - Plus Affiliation
abstract: |
  To create a reproducible analysis, it is not enough to write clean, well-documented code. 
  One must also ensure that all software dependencies are clearly declared and, ideally, 
  provide a simple way to install them. Several tools exist for this purpose, such as the 
  containerization solution [Docker]{.proglang}. A complete solution, however, requires not only 
  managing the computational environment but also orchestrating the execution of the analysis itself
  in a robust and automated way. This paper presents a comprehensive framework based on the [Nix]{.pkg} 
  package manager to solve both problems. We first introduce the [R]{.proglang} package [rix]{.pkg}, 
  which simplifies the creation of reproducible environments with [Nix]{.pkg}. We then introduce 
  [rixpress]{.pkg}, a second [R]{.proglang} package that leverages these environments to define and 
  execute complex, multi-language analytical pipelines. Together, these tools provide a powerful, 
  unified approach to achieving high standards of computational reproducibility.

@@

A final step towards the gold standard is to use a build
automation tool such as [Make]{.pkg} to run the whole analysis when the
container is executed, instead of executing scripts. Build automation tools make
it easier to run arbitrary code in a series of well-defined steps.
[R]{.proglang} programmers can use the [targets]{.pkg} by @landau2021 as a build
automation tool.

@@

However, reaching this gold standard is quite costly: one needs to learn a tool
to deal with package versions, then [Docker]{.pkg} for all the other software,
including the programming language that was used, as well as a build automation
tool. It should also be noted
that these requirements are not limited to the [R]{.proglang} programming
language. A very similar approach should be taken if one uses
[Python]{.proglang}, or any other programming language. Furthermore, the
previous discussion also implies that for an analysis to actually be
reproducible at all, it must be made using free and open source software, as
proprietary software cannot be freely shared with others that wish to reproduce
a study, and there's no guarantee that others will have access to the
proprietary software used.

As an alternative to the above approach, we will present the [Nix]{.pkg} package
manager, which is available for all major operating systems and focuses on
installing and building software in a reproducible manner. This means that this
single tool can replace not only [Docker]{.pkg} but also [renv]{.pkg} and even a
build automation tool such as [Make]{.pkg}. To make [Nix]{.pkg} more accessible
to [R]{.proglang} programmers, we developed the [rix]{.pkg} and [rixpress]{.pkgs} packages which we
will present in this article.

@@

## Orchestrating the Workflow with `rixpress` {#sec-rixpress}

Defining a reproducible environment with [rix]{.pkg} solves the first major challenge of reproducibility. The second is to reliably and efficiently execute the analysis workflow within that environment. This is the role of the [rixpress]{.pkg} package.

As mentioned in the introduction, a build automation tool like [targets]{.pkg} is invaluable for managing complex analyses. It automatically tracks dependencies between code and data objects, caching results and only re-computing steps that have changed. One can run a [targets]{.pkg} pipeline inside a Nix environment to make it reproducible. However, this approach has limitations: the entire pipeline must run in a single environment, and orchestrating steps across different languages (e.g., [R]{.proglang} and [Python]{.proglang}) requires manual handling via packages like [reticulate]{.pkg}.

[rixpress]{.pkg} overcomes these limitations by using Nix not just as a package manager, but as the build automation engine itself. In a [rixpress]{.pkg} pipeline, each step is defined as a Nix derivation. This powerful abstraction provides two key benefits:

1.  True Polyglot Pipelines: Each step can be assigned its own, distinct Nix environment. A Python step can run in a pure Python environment, an [R]{.proglang} step in an [R]{.proglang} environment, and a Quarto rendering step in yet another, all within the same pipeline.

2.  Deep Reproducibility: Because each step is a Nix derivation, its execution is hermetically sealed and its output is cached in the Nix store based on the hash of all its inputs. This provides caching and reproducibility at a level deeper than most other tools. On top of this, all artifacts built by the pipeline are direct children of the computational environment: should the environment change in any way, all the artifacts would get rebuilt, ensuring that the environment and the artifacts are never out of sync.

The user defines the pipeline in a simple [R]{.proglang} script, using functions heavily inspired by [targets]{.pkg}. Let's consider a simple pipeline that reads data with Python (`polars`), filters it, passes it to [R]{.proglang} (`dplyr`) for further analysis, and then compiles a Quarto document.

First, the pipeline is defined in an [R]{.proglang} script as a list of `rxp_*` function calls:

```{r, eval=FALSE}
# In a file like 'pipeline.R'
library(rixpress)

# Define the sequence of steps
pipeline_steps <- list(
  # Step 1: Read data using Python in its own environment
  rxp_py_file(
    name = mtcars_pl,
    path = "data/mtcars.csv",
    read_function = "lambda x: polars.read_csv(x, separator='|')",
    nix_env = "python-env.nix" # A Nix file defining the Python env
  ),

  # Step 2: Filter the data in Python
  rxp_py(
    name = mtcars_filtered,
    expr = "mtcars_pl.filter(polars.col('am') == 1).to_pandas()",
    nix_env = "python-env.nix"
  ),

  # Step 3: Convert the Python pandas DataFrame to an R data.frame
  # This uses the main project environment ('default.nix')
  rxp_py2r(
    name = mtcars_r,
    expr = mtcars_filtered
  ),

  # Step 4: Process the data in R
  rxp_r(
    name = mtcars_summary,
    expr = dplyr::summarise(mtcars_r, avg_mpg = mean(mpg))
  )
)

# Generate the 'pipeline.nix' file from the R list
rixpress(pipeline_steps, project_path = ".")
```

The user first creates the necessary environment files (e.g., `default.nix`, `python-env.nix`) using `rix`. 
Then, running the script above does not execute the pipeline but instead generates a `pipeline.nix` file, which is the complete, declarative definition of the workflow.

To build the pipeline, the user runs:

```{r, eval=FALSE}
rxp_make()
```

[Nix]{.pkg} then executes each step in the correct order, building any dependencies as needed. Because the output of each step is cached, subsequent runs of `rxp_make()` will only re-execute steps whose code or inputs have changed. This provides the same efficiency as [targets]{.pkg} but with the added benefits of multi-language support and the bit-for-bit reproducibility of Nix. Artifacts from any step can be loaded into an interactive [R]{.proglang} session for inspection using `rxp_read("artifact_name")` or `rxp_load("artifact_name")`.

For Python users, there is also a port of [rixpress]{.pkg}, called [ryxpress]{.pkg} that uses the same definition of the pipeline as an R script, but allows building the pipeline and viewing the outputs from a Python session.
